Prelim Outline

### 2018-09-19 prelim/draft.tex
Prelim draft intro looking good, need to do last 2 paragraphs;

\begin{document}
\section{Introduction} 
Understanding the low energy behavior of the cuprate superconductors has been an open question in theoretical condensed matter physics for over thirty years. 
The cuprate materials all share a similar structure of repeating layers of square lattice CuO$_2$ planes and charge reservoir blocks, of which carriers on the prior exhibit the most interesting collective behavior. 
At zero doping the cuprates are antiferromagnetic charge-transfer insulators \cite{RevModPhys.78.17}. 
The theoretical challenge arises away from zero doping where the poorly understood superconducting, pseudo-gap, strange metal phases appear \cite{RevModPhys.84.1383}. 
\ref{fig1} shows the phase diagram of a hole- and electron-doped cuprate materials with different structures. 
The superconducting phase is d-wave in character and cannot be described by phonon-induced pairings \cite{PhysRevLett.71.2134,PhysRevLett.69.1431,PhysRevB.43.2778, PhysRevB.47.11314, PhysRevLett.70.1553, RevModPhys.67.515, RevModPhys.77.109, RevModPhys.79.353, RevModPhys.75.473}. 
In the pseudo-gap phase the d-wave gap from the superconducting state continues to partially exist, so that parts of the Fermi surface are gapped above the superconducting transition temperature \cite{ROSSATMIGNOD199186, RevModPhys.77.721, PhysRevLett.75.4114, PhysRevB.61.9752}.
The strange metal phase in the hole doped cuprates is metallic but has a "strange" resistivity which does not depend quadratically on temperature like normal Fermi liquids \cite{PhysRevLett.69.2975}, but does have T-quadratic resitivity for the electron-doped materials \cite{1989PhyC..161..415T}.

The strong coupling between magnetic, charge, and lattice degrees of freedom in the cuprates makes writing down an accurate low energy effective theory very challenging. 
For example, in La$_{2-x}$Ba$_x$CuO$_4$ (LBCO) there is a structural phase transition as doping and temperature are varied between a high temperature tetragonal phase (HTT), a low-temperature orthorhombic phase (LTO) and a low temperature tetragonal phase (LTT), and the fraction of the bulk transformed material from LTO to LTT is correlated with the bulk superconducting temperature under various dopings \cite{PhysRevLett.62.2751}. 
An independent study shows that in La$_{2-x}$Sr$_x$CuO$_{4-\delta}$ (LSCO) the superconducting phase transition for x $<$ 0.19 is accompanied by a structural phase transition from the HTT to LTO phase \cite{PhysRevB.35.7191}. 
Further, there is evidence for static stripe phase, where antiferromagnetic stripes of copper spins are separated by periodically spaced domain walls to which the holes segregate, in La$_{1.6-x}$Nd$_{0.4}$Sr$_x$CuO$_4$ at x=0.12 which is associated with an anomalous suppression in the superconducting temperature \cite{Tranquada1995, PhysRevB.54.7489}. 
Elastic magnetic measurements on La$_2$CuO$_{4+y}$ show the superconducting phase is accompanied by a spin-density wave emergence indicating a strong correlation between the two phases \cite{PhysRevB.60.3643}. 
In the undoped phase of the cuprates the magnetic dispersion across the magnetic Brillouin zone strongly resembles that of spin-waves in a square Heisenberg antiferromagnet \cite{PhysRevLett.104.077002, PhysRevLett.102.167401, PhysRevLett.105.157006}, however within the superconducting phase the magnetic dispersion takes on an "hourglass" shape which has been reproduced in experiments for LSCO at x=0.10, 0.16 \cite{PhysRevLett.93.147002}, LBCO at x=0.125 \cite{Tranquada2004}, and YBa$_2$Cu$_3$O$_{6+x}$ at x=0.5, 0.6 \cite{PhysRevB.71.024522, Hayden2004}.

Many theoretical approaches to generating a low energy model for the cuprates which can account for the strong correlations between magnetic, lattice and charge degrees of freedom have been attempted with varying degrees of success. 
For the undoped cuprates a nearest neighbor Heisenberg model can reproduce the magnetic dispersions measured in inelastic neutron scattering experiments accurate away from the magnetic Brillouin zone boundary, but fail to capture the small effects of anomalous dispersion near the boundary \cite{PhysRevB.79.155114, PhysRevLett.75.553}. 
The simplest model for the hole-doped cuprates was suggested by Anderson and is a 2D single-band Hubbard model which only includes bands of Cu d$_{x^2-y^2}$ character, which will be hybridized with the oxygen bands to some degree. 
In the strong coupling limit at half filling and T=0 this model reduces to a nearest neighbor Heisenberg model with long-range AFM order \cite{doi:10.1139/p78-120, PhysRevB.38.316}.
The model even exhibits a pseudo-gap like phase in the underdoped region \cite{PhysRevLett.94.156404,PhysRevB.74.024508}. 
Further, numerical studies on the Hubbard model show the existence of a $d_{x^2-y^2}$ superconducting-like phase \cite{PhysRevB.81.224505, PhysRevB.70.054504}. 
A similar model is the t-J model  which is built on the idea that under doping there are itinerant holes which hop around the oxygen bands alongside the fixed S=1/2 moments on the copper sites which interact through a superexchange term with parameter J. 
The t-J model can be seen as an effective theory of the single band t-U model. 
This model exhibits superconductivity at J/t $\sim$ 3 with a density of $\langle n \rangle$ = 1/2 \cite{PhysRevB.45.5744, PhysRevLett.70.682}.
While these models have fewer degrees of freedom and thus allow for quick numerical computations, it has been shown in accurate \textit{ab initio} Quantum Monte Carlo calculations that holes move to the oxygen bands under doping, indicating that an effective theory should include the oxygen bands explicitly not implicity inclusion via hybridization. 
This leads to a 2D 3-band Hubbard model which includes the Cu 3$d_{x^2-y^2}$ and O p bands \cite{PhysRevLett.58.2794, VARMA1987681} and exhibits qualitatively similar behavior to the 1-band model. 
However, these approaches to generating a low-energy model have two inherent flaws. 
First, the model parameters are not specified with the model, meaning that seeing whether the model can accurately describe the cuprates involves mapping the entire parameter space. 
Second, there is no standard measure of effectiveness of the models under question: some cherry-picked qualitative and quantiative features of the model are compared to reality.

We propose that a density matrix downfolding (DMD) \cite{10.3389/fphy.2018.00043, doi:10.1063/1.4927664} using \textit{ab initio} quantum Monte Carlo calculations (QMC) can generate an effective model which can accurately describe low-energy properties and energetics for the infinite layer electron-doped cuprates at zero temperature and dopings of x=0, 0.125, 0.25. 
We choose to study the electron-doped cuprates because while they do not exhibit the exact same features as the hole-doped materials, they do have AFM,  superconducting, and "normal" metallic phases. 
Further, the electron-doped materials have fewer relevant degrees of freedom to worry, as electrons added to the x=0 CuO$_2$ plane can only fall into the unoccupied Cu 3d orbitals, allowing us to ignore the explicit inclusion of O p bands in our study. 
The choice of x=0, 0.125 and 0.25 at T=0 allows us to probe the most robust parts of the AFM, superconducting and metallic phases respectively, ignoring any thermal fluctuations or phase transitions like the LTO to LTT transition described earlier. 
The DMD downfolding approach allows us to map in a systematic way the \textit{ab initio} Hamiltonian to an effective model Hamiltonian, with specified model parameters and basis elements, in a way which ensures the model accurately reproduces the expectation value of the \textit{ab initio} Hamiltonian for states in a low-energy subspace of the full Hilbert space. 
In order to conduct the DMD procedure we require states in the low-energy subspace of the full \textit{ab initio} Hilbert space. 
These states will be sampled using fixed-node diffusion Monte Carlo, a QMC algorithm which can project out low-energy states starting with appropriate trial wave functions \cite{RevModPhys.73.33}.

The density matrix downfolding method (DMD) allows us to downfold a first-principles Hamiltonian into a low-energy effective theory by mapping the downfolding problem to a linear regression problem which uses ab-initio energies and reduced density matrices of wave functions sampled from a low-energy subspace of the full Hilbert space \cite{10.3389/fphy.2018.00043}. 

@@

We will use fixed node diffusion Monte Carlo (FN-DMC) to generate
wave functions in our chosen low-energy subspace and to accurately calculate ab-initio energies and reduced density matrices. 
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a Hamiltonian given some initial trial wave function. 
Consider a trial wave function $|\psi\rangle$ and a Hamiltonian $H$ with ground state $|\phi_0\rangle$. 
We apply the projector $e^{-\tau (H-E_0)}$ as $\tau \rightarrow \infty$ to $|\psi \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau (H-E_0)} |\psi\rangle \propto \langle \phi_0|\psi\rangle |\phi_0\rangle,
\end{equation}
projecting out the ground state $|\phi_0\rangle$ of $H$ as long as the trial wave function we choose is not orthogonal to the ground state. 
The stochastic implementation involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle$. 
Since the Hamiltonian has two parts, $H = T + V$ a Trotter expansion is used to approximate the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle \sim e^{-\tau(V(R) + V(R^\prime))/2} \langle R| e^{-\tau(T - E_T)}|R^\prime \rangle + O(\tau^2) $ which leads to a time step error in the calculation if the time step between moves $\tau$ is too large. 
This implementation when used on a Hamiltonian with fermions also leads to a fermion sign problem. 
We deal with this sign problem through a fixed-node approximation, where the nodal surface of the projected wave function is forced to match that of the initial trial wave function. 
This approximation makes FN-DMC variational, and will only return the exact ground state of $H$ if the nodal surfaces of $|\psi\rangle$ and $|\phi_0\rangle$ are identical. 
Therefore the FN-DMC calculation has two major systematic errors that have to be dealt with, the fixed-node error and the time-step error. 
The time-step error is typically handled through a time-step extrapolation, where multiple FN-DMC calculations are conducted at sequentially smaller $\tau$ and the relevant expectation values are extrapolated to $\tau \rightarrow 0$ \cite{Needs2010}. 
Fixed-node error can be reduced by using optimized many-body trial wave functions of the (multi-)Slater-Jastrow form seen in \cite{PhysRevLett.98.110201}.

>>>>

\begin{document}
\section{Introduction} 
\textbf{Understanding the low energy behavior of the cuprate superconductors has been an open question in theoretical condensed matter physics for over thirty years.}
The cuprate materials all share a similar structure of repeating layers of square lattice CuO$_2$ planes and charge reservoir blocks, of which carriers on the prior exhibit the most interesting collective behavior. 
At zero doping the cuprates are antiferromagnetic charge-transfer insulators \cite{RevModPhys.78.17}. 
The theoretical challenge arises away from zero doping where the poorly understood superconducting, pseudo-gap, and strange metal phases appear \cite{RevModPhys.84.1383}. 
\ref{fig1} shows the phase diagram of hole- and electron-doped cuprate materials with different structures. 
The superconducting phase is d-wave in character and cannot be described by phonon-induced pairings \cite{PhysRevLett.71.2134,PhysRevLett.69.1431,PhysRevB.43.2778, PhysRevB.47.11314, PhysRevLett.70.1553, RevModPhys.67.515, RevModPhys.77.109, RevModPhys.79.353, RevModPhys.75.473}. 
In the pseudo-gap phase the d-wave gap from the superconducting state continues to partially exist, so that parts of the Fermi surface are gapped above the superconducting transition temperature \cite{ROSSATMIGNOD199186, RevModPhys.77.721, PhysRevLett.75.4114, PhysRevB.61.9752}.
The strange metal phase in the hole doped cuprates is metallic but has a "strange" resistivity which does not depend quadratically on temperature like normal Fermi liquids \cite{PhysRevLett.69.2975}, but does have T-quadratic resitivity for the electron-doped materials \cite{1989PhyC..161..415T}.

\textbf{The strong coupling between magnetic, charge, and lattice degrees of freedom in the cuprates makes writing down a compact and accurate low energy effective theory which can describe the breadth of phenomenon related to these degrees of freedom very challenging.}
As an example of strong coupling, consider the structural phase transitions in cuprates which accompany the superconducting transition.
In La$_{2-x}$Ba$_x$CuO$_4$ (LBCO) structural phase transitions occur as doping and temperature are varied between a high temperature tetragonal phase (HTT), a low-temperature orthorhombic phase (LTO) and a low temperature tetragonal phase (LTT), and the fraction of the bulk transformed material from LTO to LTT is correlated with the bulk superconducting temperature under various dopings \cite{PhysRevLett.62.2751}. 
An independent study shows that in La$_{2-x}$Sr$_x$CuO$_{4-\delta}$ (LSCO) the superconducting phase transition for x $<$ 0.19 is accompanied by a structural phase transition from the HTT to LTO phase \cite{PhysRevB.35.7191}. 
Similarly, changes in magnetic phases and properties appear to be correlated with the superconducting phase, again providing evidence for strong coupling of microscopic degrees of freedom.
For example, there is evidence for a static stripe phase, where antiferromagnetic stripes of copper spins are separated by periodically spaced domain walls to which the holes segregate, in La$_{1.6-x}$Nd$_{0.4}$Sr$_x$CuO$_4$ at x=0.12 which is associated with an anomalous suppression in the superconducting temperature \cite{Tranquada1995, PhysRevB.54.7489}. 
Elastic magnetic measurements on La$_2$CuO$_{4+y}$ show the superconducting phase is accompanied by a spin-density wave emergence indicating a strong correlation between the two phases \cite{PhysRevB.60.3643}. 
In the undoped phase of the cuprates the magnetic dispersion across the magnetic Brillouin zone strongly resembles that of spin-waves in a square Heisenberg antiferromagnet \cite{PhysRevLett.104.077002, PhysRevLett.102.167401, PhysRevLett.105.157006}, however within the superconducting phase the magnetic dispersion takes on an "hourglass" shape which has been reproduced in experiments for LSCO at x=0.10, 0.16 \cite{PhysRevLett.93.147002}, LBCO at x=0.125 \cite{Tranquada2004}, and YBa$_2$Cu$_3$O$_{6+x}$ at x=0.5, 0.6 \cite{PhysRevB.71.024522, Hayden2004}. 
These experiments indicate that in order to write down a low-energy model for the cuprates, one has to generate effective degrees of freedom and interactions which are complicated combinations of the microscopic lattice, charge, and spin degrees of freedom.
The strong coupling between the microscopic degrees of freedom indicate that these combinations cannot be approached in a perturbative fashion from either the weak or very-strong coupling limit.

\textbf{Various models have been proposed and studied as candidates for effective theories of the cuprates, however these models were neither developed nor validated in a systematic way.}
For the undoped cuprates a nearest neighbor Heisenberg model can reproduce the magnetic dispersions measured in inelastic neutron scattering experiments accurately away from the magnetic Brillouin zone boundary, but fail to capture the small effects of anomalous dispersion near the boundary \cite{PhysRevB.79.155114, PhysRevLett.75.553}. 
The simplest model for the hole-doped cuprates was suggested by Anderson and is a 2D single-band Hubbard model which only includes bands of Cu d$_{x^2-y^2}$ character, which will be hybridized with the oxygen bands to some degree. 
In the strong coupling limit at half filling and T=0 this model reduces to a nearest neighbor Heisenberg model with long-range AFM order \cite{doi:10.1139/p78-120, PhysRevB.38.316}.
The model even exhibits a pseudo-gap like phase in the underdoped region \cite{PhysRevLett.94.156404,PhysRevB.74.024508}. 
Further, numerical studies on the Hubbard model show the existence of a $d_{x^2-y^2}$ superconducting-like phase \cite{PhysRevB.81.224505, PhysRevB.70.054504}. 
A similar model is the t-J model  which is built on the idea that under doping there are itinerant holes which hop around the oxygen bands alongside the fixed S=1/2 moments on the copper sites which interact through a superexchange term with parameter J. 
The t-J model can be seen as an effective theory of the single band t-U model. 
This model exhibits superconductivity at J/t $\sim$ 3 with a density of $\langle n \rangle$ = 1/2 \cite{PhysRevB.45.5744, PhysRevLett.70.682}.
While these models have fewer degrees of freedom and thus allow for quick numerical computations, it has been shown in accurate \textit{ab initio} Quantum Monte Carlo calculations that holes move to the oxygen bands under doping, indicating that an effective theory should include the oxygen bands explicitly, not implicity via hybridization. 
This leads to a 2D 3-band Hubbard model which includes the Cu 3$d_{x^2-y^2}$ and O p bands \cite{PhysRevLett.58.2794, VARMA1987681} and exhibits qualitatively similar behavior to the 1-band model. 
However, these approaches to generating a low-energy model have multiple issues.
First, generating the model - specifically the effective degrees of freedom, interactions and model parameters - was not done systematically, rather the choice of effective model was \textit{ad hoc}.
Second, there is no standard measure of effectiveness of the models under question. 
Typically, just a few qualitative and quantitative features of the model are compared to reality as a form of validation.

\textbf{We propose that a density matrix downfolding (DMD) \cite{10.3389/fphy.2018.00043, doi:10.1063/1.4927664} technique using \textit{ab initio} quantum Monte Carlo calculations (QMC) can systematically generate an effective model that accurately describes the low-energy properties of cuprate materials, which we will test on the electron-doped cuprate SrCuO$_2$ at T=0 and x=0, 0.125, 0.25. }
We choose to study the electron-doped cuprates because while they do not exhibit the exact same features as the hole-doped materials, they do have AFM,  superconducting, and pseudo-gap phases.
Further, the electron-doped materials have fewer relevant degrees of freedom, as electrons added to the x=0 CuO$_2$ plane can only fall into the unoccupied Cu 3d orbitals, allowing us to ignore the explicit inclusion of O p bands in our study. 
The choice of x=0, 0.125 and 0.25 at T=0 allows us to probe the most robust parts of the AFM, superconducting and metallic phases respectively, ignoring any thermal fluctuations or phase transitions like the LTO to LTT transition described earlier. 
In order to reduce the computational cost we look at the infinite-layer materials which have unit cells with the fewest number of electrons within the cuprate family.
The DMD approach allows us to map in a systematic way the \textit{ab initio} Hamiltonian to an effective model Hamiltonian, with specified model parameters and basis elements, while ensuring the model accurately reproduces the expectation value of the \textit{ab initio} Hamiltonian for states in a low-energy subspace of the full Hilbert space. 
The DMD approach also has a validation procedure which allows us to assess the effectiveness of the low-energy model in a standardized way.
In order to conduct the DMD procedure we require states in the low-energy subspace of the full \textit{ab initio} Hilbert space, which will be sampled using fixed-node diffusion Monte Carlo, a QMC algorithm which can project out low-energy states starting with appropriate trial wave functions \cite{RevModPhys.73.33}.

\textbf{The density matrix downfolding method (DMD) allows us to downfold a first-principles Hamiltonian into a low-energy effective theory by mapping the downfolding problem to a linear regression problem which uses ab-initio energies and reduced density matrices of wave functions sampled from a low-energy subspace of the full Hilbert space \cite{10.3389/fphy.2018.00043}. }

@@

\textbf{We will use fixed node diffusion Monte Carlo (FN-DMC) to generate
wave functions in our chosen low-energy subspace and to accurately calculate ab-initio energies and reduced density matrices. }
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a Hamiltonian given some initial trial wave function. 
Consider a trial wave function $|\psi\rangle$ and a Hamiltonian $H$ with ground state $|\phi_0\rangle$. 
We apply the projector $e^{-\tau (H-E_0)}$ as $\tau \rightarrow \infty$ to $|\psi \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau (H-E_0)} |\psi\rangle \propto \langle \phi_0|\psi\rangle |\phi_0\rangle,
\end{equation}
projecting out the ground state $|\phi_0\rangle$ of $H$ as long as the trial wave function we choose is not orthogonal to the ground state. 
The stochastic implementation involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle$. 
Since the Hamiltonian has two parts, $H = T + V$ a Trotter expansion is used to approximate the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle \sim e^{-\tau(V(R) + V(R^\prime))/2} \langle R| e^{-\tau(T - E_T)}|R^\prime \rangle + O(\tau^2) $ which leads to a time step error in the calculation if the time step between moves $\tau$ is too large. 
This implementation when used on a Hamiltonian with fermions also leads to a fermion sign problem. 
We deal with this sign problem through a fixed-node approximation, where the nodal surface of the projected wave function is forced to match that of the initial trial wave function. 
This approximation makes FN-DMC variational, and will only return the exact ground state of $H$ if the nodal surfaces of $|\psi\rangle$ and $|\phi_0\rangle$ are identical. 
Therefore the FN-DMC calculation has two major systematic errors that have to be dealt with, the fixed-node error and the time-step error. 
The time-step error is typically handled through a time-step extrapolation, where multiple FN-DMC calculations are conducted at sequentially smaller $\tau$ and the relevant expectation values are extrapolated to $\tau \rightarrow 0$ \cite{Needs2010}. 
Fixed-node error can be reduced by using optimized many-body trial wave functions of the (multi-)Slater-Jastrow form seen in \cite{PhysRevLett.98.110201}.
